---
title: "Correlating Like a Data Master"
author: "Abhishek Deore"
format: html
editor: visual
---

## Required Set-up

Initially, we will prepare our environment with the necessary packages.

```{r, warning=FALSE}
if (!require(pacman))
  install.packages("pacman")

pacman::p_load(colorblindr,
       dlookr,
       formattable,
       GGally,
       ggdist,
       ggpubr,
       ggridges,
       here,
       tidyverse)

# Set global ggplot() theme
# Theme pub_clean() from the ggpubr package with base text size = 16
theme_set(theme_pubclean(base_size = 12)) 
# All axes titles to their respective far right sides
theme_update(axis.title = element_text(hjust = 1))
# Remove axes ticks
theme_update(axis.ticks = element_blank()) 
# Remove legend key
theme_update(legend.key = element_blank())
```

## Load and Examine the Data-set

```{r}
artists <- readr::read_csv('https://raw.githubusercontent.com/rfordatascience/tidytuesday/master/data/2023/2023-01-17/artists.csv')



artists |>
  head()|>
  formattable()

```

## **Describe and Visualize Correlations**

Correlations represent statistical associations between two numeric variables, and they may or may not imply causation. Investigating correlations within your data is essential for assessing data independence, a crucial assumption for parametric statistical methods, ensuring that your variables were collected randomly.

This exploration helps identify potential relationships and dependencies among variables, aiding in data analysis and hypothesis testing.

```{r}
# Table of correlations between numerical variables (we are sticking to the default Pearson's r coefficient)
correlate(artists) |>
  formattable()
```

```{r}
# Correlation matrix of numerical variables
artists |>
  plot_correlate()
```

This is a correlation matrix plot for numerical variables which shows the relationship between variables through a heatmap like graph.

## **Visualize Correlations within Groups**

It is always advisable to examine the correlations among numerical variables within specific groups if we plan to compare these groups later on. This preliminary analysis helps understand how variables behave within distinct categories and can provide insights for subsequent comparisons.

```{r, warning=FALSE}
artists |>
  group_by(artist_nationality) |>
  plot_correlate()
```

Above is the correlation Matrix of Numerical Variables grouped by Artist Nationality.

We will utilize the **`ggpairs()`** function from the **`GGally`** package to examine the correlations between (**`space_ratio_per_page_total`**) and (**`artist_unique_id`**).

```{r}
artists |> 
  dplyr::select(book, space_ratio_per_page_total, artist_unique_id) |>
  ggpairs(aes(color = book, alpha = 0.5)) +
  theme(strip.background = element_blank())
  
```

Above plot displays relationships between '**`space_ratio_per_page_total`**' and '**`artist_unique_id`**' variables, colored by '**`book`**' with transparency, emphasizing potential correlations and patterns.

## **Describe and Visualize Relationships Based on Target Variables**

### Target Variables

`Target variables` are essentially numerical or categorical variables that you want to relate others to in a data frame. `dlookr` does this through the `target_by()` function, which is similar to `group_by()` in `dplyr`. The `relate()` function then briefly analyzes the relationship between the target variable and the variables of interest.

The relationships below will have the formula relationship `target ~ predictor`.

### **Numerical Target Variables: Numerical Variable of Interest**

```{r}
# First, we need to remove NAs, they cause an error
dataset.noNA <- artists |> 
  drop_na()

# The numerical predictor variable that we want
num <- target_by(dataset.noNA, edition_number)

# Relating the variable of interest to the numerical target variable
num_num <- relate(num, space_ratio_per_page_total)

# Summary of the regression analysis - the same as the summary from lm(Formula)
summary(num_num)
```

```{r}
# Plotting the linear relationship
plot(num_num)
```

Above plot depicts the linear relationship between '**`edition_number`**' and '**`space_ratio_per_page_total`**' variables after removing NA values and conducting a regression analysis.

### **Numerical Target Variables: Categorical Variable of Interest**

```{r}
# The categorical predictor variable that we want
num <- target_by(artists, space_ratio_per_page_total) 

# We need to change Group to a factor
num$book <- as.factor(num$book)

# Relating the variable of interest to the numerical target variable
num_cat <- relate(num, book)

# Summary of the ANOVA analysis - the same as the summary from anova(lm(Formula))
summary(num_cat)
```

```{r}
plot(num_cat) + 
  theme(axis.text.x = element_blank())
```

Above is the box plot illustrating the results of an ANOVA analysis, examining the relationship between the categorical variable '**`book`**' and the numerical variable '**`space_ratio_per_page_total`**'.

### **Categorical Target Variables: Numerical Variable of Interest**

Note that this produces descriptive statistics, unlike the other relationships we are looking at.

```{r, warning=FALSE}
# The categorical predictor variable that we want
categ <- target_by(artists, book)

# Relating the variable of interest to the numerical target variable
cat_num <- relate(categ, space_ratio_per_page_total)

# Summary of descriptive statistics
summary(cat_num)
```

```{r}
plot(cat_num) 
```

Above is the density plot showcasing the distribution of '**`space_ratio_per_page_total`**' within different '**`book`**' categories

### **Categorical Target Variables: Categorical Variable of Interest**

```{r, warning=FALSE}
# The categorical predictor variable that we want
categ <- target_by(artists, artist_gender)

# Relating the variable of interest to the categorical target variable
cat_cat <- relate(categ, book)

# Summary of the Chi-square test for Independence
summary(cat_cat)
```

```{r}
plot(cat_cat)
```

Above is the clustered bar plot depicting the results of a Chi-square test for Independence, examining the relationship between the categorical variable '**`book`**' and the categorical variable '**`artist_gender`**'
